---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Loading relevant libraries

```{r}
library(tidyverse)
library(doBy)
```

Loading and preprocessing the data

```{r}
setwd("/Users/jesus/Documents/GitHub/RepData_PeerAssessment1/activity")
RawData<- read.csv("activity.csv",na.strings=("NA"))

```

## What is mean total number of steps taken per day?

```{r}
NaRemoved<-filter(RawData, (is.na(RawData$steps)==FALSE))
StepsPerDay<-summaryBy(steps~date,data=NaRemoved, FUN = sum)
ggplot(StepsPerDay, aes(x=steps.sum)) + geom_histogram()+xlab("Steps")+ylab("Days")

MeanSteps<-mean(StepsPerDay$steps.sum)
MedSteps<-median(StepsPerDay$steps.sum)

print(c("Mean steps taken is",MeanSteps))
print(c("Median steps taken is",MedSteps))

```


## What is the average daily activity pattern?
```{r, echo=TRUE}
StepsPerInterval<-summaryBy(steps~interval,data=NaRemoved, FUN = mean)

# Converts from format hhmm to minutes
ConvertedSteps<-mutate(StepsPerInterval,interval = str_pad(interval,

                            width = 4,

                            side = "left",

                            pad = "0")) %>%

  extract(interval,

          into = c("hr", "min"),

          regex = "(\\d{2})(\\d{2})",

          convert = TRUE) %>%

  mutate(interval = (hr*60) + (min))

#graph of steps over the interval using minutes
ggplot(ConvertedSteps, aes(x=interval, y=steps.mean) )+geom_line()+xlab("Minutes")+ylab("steps")

#Calculation of which interval had the max steps taken using t
StepsPerInterval$interval[which.max(StepsPerInterval$steps.mean)]


```
The interval with the max steps taken is reported by the code above as 835, indicating 8hours and 35 minutes.

## Imputing missing values

```{r}
#Calculation of number of missing values
print(nrow(RawData[is.na(RawData$steps)==TRUE,]))

#Calculation of Steps/Interval of non-NA data
RawStepsPerInterval <- summaryBy(steps~interval,data=NaRemoved, FUN = mean)

#Creation of Imputed Data set
MergedResults<-merge(RawData,RawStepsPerInterval, by.x = "interval", by.y = "interval", all.x=TRUE)
MergedResults$ImputedSteps<-coalesce(MergedResults$steps,MergedResults$steps.mean)

ImputedData<-summaryBy(ImputedSteps~date,data=MergedResults, FUN = sum)

#Histogram plot of Imputed Data Set Steps per Day
ggplot(ImputedData, aes(x=ImputedSteps.sum))+geom_histogram()+xlab("Steps")+ylab("Days")

ImputedMeanSteps<-mean(ImputedData$ImputedSteps.sum)
ImputedMedSteps<-median(ImputedData$ImputedSteps.sum)

print(c("Mean steps taken is",ImputedMeanSteps))
print(c("Median steps taken is",ImputedMedSteps))

```
By imputing the data, we see that the shape of the histogram does not change drastically indicating minimal change of estimates.The mean and the median become equal.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
WeekendData<-MergedResults %>%

  mutate(date=as.POSIXct(date), Day=weekdays(date)) %>%


  mutate(DayType= if_else(Day=="Saturday"| Day=="Sunday","weekend","weekday") ) %>%

mutate(intervalMin = str_pad(interval,

                            width = 4,

                            side = "left",

                            pad = "0")) %>%

  extract(intervalMin,

          into = c("hr", "min"),

          regex = "(\\d{2})(\\d{2})",

          convert = TRUE) %>%

  mutate(intervalMin = (hr*60) + (min)) %>%
  group_by(DayType, intervalMin) %>%
  summarize(MeanSteps=mean(ImputedSteps))

ggplot(WeekendData, aes(x=intervalMin, y=MeanSteps) )+geom_line()+xlab("Minutes")+ylab("steps")+facet_wrap(~DayType,ncol = 1)

```



Activity starts earlier on weekdays, has a greater step/interval maximum, but ends earlier. Activity is spread throughout the weekend.